---
title: Suicide risk communication and intervention preferences for veterans and service
  members.
output:
  pdf_document: default
  html_document: default
date: "2023-08-14"
---

## Install Packages and Load Dataset
```{r, results=F, warning=F, message=F}
library(dataMaid)
library(lubridate)
library(summarytools)
library(ggstatsplot)
library(ggplot2)
library(gridExtra)
library(tidyr)
library(Rmisc)
library(ltm)
library(ggfortify)
library(ggpubr)
library(factoextra)
library(cluster)
library(nlme)
library(emmeans)
library(dplyr)
library(ggpubr)
```

```{r, results=F, warning=F, message=F}
#load data set
vettix<-read.csv("SSS_Poll_wVAQuestions.csv")
```

## Clean and Set Variables
```{r, results=F, warning=F, message=F}
#Filter data to remove respondents that were never in the military, LEO reported branch, or Space Force
  #Space Force participants excluded due to small sample size and potential for identifiable data
#filter data to keep only veterans and currently serving participants
#Mutate data to create a dataframe counting all participants that took part in the SSS survey
#Mutate data to create a dataframe counting all participants that took part in the VA survey
table(vettix$Branch)

vettix<-vettix%>%
  filter(Branch != "I was never in the service" & Branch != "LEO" & Branch != "Space Force")%>%
  filter(Status == "I am a Severely Wounded Veteran" | Status == "I am a Veteran" | Status == "I am Currently Serving")%>%
  mutate(SSS.Survey.Response=(Who.Family=="No"|Who.Family=="Yes"|How.Social.Media.General=="No"|How.Social.Media.General=="Yes"|Allow.Knowing=="No"|Allow.Knowing=="Yes"|Suicide.Thoughts=="No"|Suicide.Thoughts=="Yes"|Suicide.Thoughts=="I prefer not to say")*1)%>%
  mutate(VA.Survey.Response=(Registered.With.VA !=""|VA.Disability.Claim=="Yes"|VA.Disability.Claim=="No")*1)

#Sanity check to ensure the filter had been successfully implemented
table(vettix$Branch)
#Get a count of observations that included those participants that answered SSS survey questions
table(vettix$SSS.Survey.Response)

#Get a count of demographic variables to see the range, and remove values which are not reasonable, replacing with an 'NA'
table(vettix$Age)
#Remove all participants with an estimated age over 100
vettix$Age[vettix$Age>100]<-NA
table(vettix$Age2)
vettix$Age2[vettix$Age2>100]<-NA
table(vettix$Yrs.Served)
#Remove all participants with more than 65 years served
vettix$Yrs.Served[vettix$Yrs.Served>65]<-NA
#Replace all empty cells with an 'NA'
vettix[vettix==""]<-NA


#Create binned age variable 
vettix$age.cat <- cut(vettix$Age2, breaks=c(18,25,35,45, 55, 65, 91), labels=c("18-24","25-34","35-44", "45-54", "55-64", "65-91"))
#Sanity check of categorization
vettix[,c(3,59)]

#Create binned years served variable
vettix$yrs.ser.cat <- cut(vettix$Yrs.Served, breaks=c(0,2,5,10,20,40,60), labels=c("0-2", "3-5", "6-10", "11-20","21-40","41-65"))
#Sanity check of categorization
vettix[,c(4,60)]

#create factor variables
vettix$Branch<-as.factor(vettix$Branch)
vettix$Status<-as.factor(vettix$Status)
vettix$Income<-as.factor(vettix$Income)
vettix$Education<-as.factor(vettix$Education)
vettix$Marital.Status<-as.factor(vettix$Marital.Status)
vettix$Highest.Rank<-as.factor(vettix$Highest.Rank)
vettix$Suicide.Thoughts[vettix$Suicide.Thoughts=="Yes"]<-"Yes."
vettix$Suicide.Thoughts[vettix$Suicide.Thoughts=="No"]<-"No."
vettix$Registered.With.VA[vettix$Registered.With.VA=="Yes"]<-"Yes."
vettix$Suicide.Thoughts<-as.factor(vettix$Suicide.Thoughts)
vettix$Registered.With.VA<-as.factor(vettix$Registered.With.VA)
table(vettix$Suicide.Thoughts)


#Create all other numeric dummy variables
  #Yes will be signified by 1 and no will be signified by 0
vettix[vettix=="Yes"]<-1
vettix[vettix=="No"]<-0

#Set all continuous variables as numeric
vettix$Who.Family<-as.numeric(vettix$Who.Family)
vettix$Who.Military.Friend<-as.numeric(vettix$Who.Military.Friend)
vettix$Who.Non.Military.Friend<-as.numeric(vettix$Who.Non.Military.Friend)
vettix$Who.Boss<-as.numeric(vettix$Who.Boss)
vettix$Who.Chaplain<-as.numeric(vettix$Who.Chaplain)
vettix$Who.MH.Provider<-as.numeric(vettix$Who.MH.Provider)
vettix$Who.Other.Provider<-as.numeric(vettix$Who.Other.Provider)
vettix$Who.VCL.SPL<-as.numeric(vettix$Who.VCL.SPL)
vettix$Who.VSO<-as.numeric(vettix$Who.VSO)
vettix$Who.No.One<-as.numeric(vettix$Who.No.One)

vettix$How.Social.Media.General<-as.numeric(vettix$How.Social.Media.General)
vettix$How.Social.Media.Suicide<-as.numeric(vettix$How.Social.Media.Suicide)
vettix$How.Online.Resources.General<-as.numeric(vettix$How.Online.Resources.General)
vettix$How.Online.Resources.Suicide<-as.numeric(vettix$How.Online.Resources.Suicide)
vettix$How.Contact.Others.General<-as.numeric(vettix$How.Contact.Others.General)
vettix$How.Contact.Others.Suicide<-as.numeric(vettix$How.Contact.Others.Suicide)
vettix$How.Others.Ask.General<-as.numeric(vettix$How.Others.Ask.General)
vettix$How.Others.Ask.Suicide<-as.numeric(vettix$How.Others.Ask.Suicide)
vettix$How.Not.At.All<-as.numeric(vettix$How.Not.At.All)

vettix$Allow.Knowing<-as.numeric(vettix$Allow.Knowing)
vettix$Allow.Lock.Up.Guns<-as.numeric(vettix$Allow.Lock.Up.Guns)
vettix$Allow.Gun.Access.Notification<-as.numeric(vettix$Allow.Gun.Access.Notification)
vettix$Allow.Holding.Guns<-as.numeric(vettix$Allow.Holding.Guns)
vettix$Allow.Providers.In.My.House<-as.numeric(vettix$Allow.Providers.In.My.House)
vettix$Allow.App.To.Help<-as.numeric(vettix$Allow.App.To.Help)
vettix$Allow.MH.Treatment<-as.numeric(vettix$Allow.MH.Treatment)
vettix$Allow.Resourcing<-as.numeric(vettix$Allow.Resourcing)
vettix$Allow.Nothing<-as.numeric(vettix$Allow.Nothing)

vettix$VA.Disability.Claim<-as.numeric(vettix$VA.Disability.Claim)
vettix$VA.Healthcare<-as.numeric(vettix$VA.Healthcare)
vettix$VA.Education<-as.numeric(vettix$VA.Education)
vettix$VA.Online.Records<-as.numeric(vettix$VA.Online.Records)
vettix$VA.Housing.Services<-as.numeric(vettix$VA.Housing.Services)
vettix$VA.Employment.Services<-as.numeric(vettix$VA.Employment.Services)
vettix$VA.Life.Insurance<-as.numeric(vettix$VA.Life.Insurance)
vettix$VA.Career.Resources<-as.numeric(vettix$VA.Career.Resources)
vettix$VA.Burial.Benefits<-as.numeric(vettix$VA.Burial.Benefits)

#create date and time variables
vettix$Who.Date<-as.POSIXct(vettix$Who.Time.Stamp, format="%m/%d/%Y %H:%M", tz="UTC")
vettix$How.Date<-as.POSIXct(vettix$How.Time.Stamp, format="%m/%d/%Y %H:%M", tz="UTC")
vettix$Allow.Date<-as.POSIXct(vettix$Allow.Time.Stamp, format="%m/%d/%Y %H:%M", tz="UTC")
vettix$Suicide.Thoughts.Date<-as.POSIXct(vettix$Suicide.Thoughts.Time.Stamp, format="%m/%d/%Y %H:%M", tz="UTC")
vettix$Registered.With.VA.Date<-as.POSIXct(vettix$Registered.With.VA.Time.Stamp, format="%m/%d/%Y %H:%M", tz="UTC")
vettix$VA.Date<-as.POSIXct(vettix$VA.Time.Stamp, format="%m/%d/%Y %H:%M", tz="UTC")
vettix<-vettix[,-c(1,22,32,42,44,46,56)]
```

## Filter for SSS Specific Survey Columns (removing VA related questions)
```{r, results=F, warning=F, message=F}
#remove columns specific to VA utilization
vettix.sss=vettix[ , c(1:39, 50, 52:55)]
#Remove all data rows that did not participate in the SSS Survey
vettix.sss=subset(vettix.sss, SSS.Survey.Response == 1)
#Use only responses in which we have a complete survey
complete.data=na.omit(vettix.sss)
```

> Total number of participants prior to removing samples that were not complete was 38,185. The complete data set, including only samples in which every variable was collected was 31,180. For use in cluster analysis and other statistical measures, we chose to proceed with the complete data set. 

## Create Demographics Summary Table
```{r, results=F, warning=F, message=F}
library(vtable)
#Print table statistics in R Markdown
st(complete.data, out='browser')

```

## Create Summary Data 
```{r, results=F, warning=F, message=F}
#Create sum totals for the number of people each participants indicted they would be willing to reach out to

complete.data=complete.data %>%
  mutate(Total.who = rowSums(across(c(11:20))))
complete.data=complete.data %>%
  mutate(Mean.who = rowMeans(across(c(11:20))))

#Create sum totals for the number of ways each participants indicted they would be willing express themselves
complete.data=complete.data %>%
  mutate(Total.how = rowSums(across(c(21:29))))
complete.data=complete.data %>%
  mutate(Mean.how = rowMeans(across(c(21:29))))

#Create sum totals for the number ways participants would allow others to intervene 

complete.data=complete.data %>%
  mutate(Total.allow = rowSums(across(c(30:38))))
complete.data=complete.data %>%
  mutate(Mean.allow = rowMeans(across(c(30:38))))

complete.data=complete.data %>%
  mutate(mean.comm = rowMeans(across(c(48,50))))
complete.data=complete.data %>%
  mutate(Total.comm = rowSums(across(c(47,49))))
```


## Calculate Proportion of Participants who chose any multi-select option with a question
```{r, warning=F}
#subset data frame to only include the columns for who/how/allow multi-select responses with additional column "Suicide.Thoughts" for grouped graphing
averages=complete.data[, c(11:39)]

average.overall=averages %>% 
  summarise_all(.funs = c(mean="mean"))
average.overall

#calculate the average percentage of participants that selected each available option by group variable "Suicide.Thoughts"
averages=averages %>% group_by(Suicide.Thoughts) %>% 
  summarise_all(.funs = c(mean="mean"))

#transition the data from wide to long format, creating a Response and Percentage column grouped by Suicide.Thoughts
averages <- gather(averages, Response, Percentage, Who.Family_mean:Allow.Nothing_mean, factor_key=TRUE)

#Remove the _mean naming scheme added to each multi-select response during the calculation process
averages$Response = gsub("_mean",'',averages$Response)
```

> Cross ref with Annerman et al, 2022
> Of these disclosures, over 87% of them were informal forms of support including friends, family, and domestic partners. The remaining 13% reported to medical professionals, mental health professionals and crisis hotlines. 

```{r, warning=F, message=F}

#Informal=Boss, Non-mil friend, Chaplain, Mil Friend, Family
#Formal=Other provider, VSO, VCL SPL, MH Provider 

dir.comp=complete.data[, c(11:39)]

dir.comp.tot=dir.comp %>% 
  summarise_all(.funs = c(mean="mean"))
dir.comp.tot

informal=mean(0.62, 0.51, 0.18, 0.04, 0.22)
formal=mean(0.44, 0.11, 0.26, 0.17)

#Who only
who.sums=colSums(dir.comp[sapply(dir.comp, is.numeric)])
who.sums=as.data.frame(who.sums)
who.sums <- tibble::rownames_to_column(who.sums, "Question")
names(who.sums)[names(who.sums) == 'who.sums'] <- 'freq'
who.inf=who.sums[c(1:5), ]
who.for=who.sums[c(6:9), ]
who.sums=who.sums[c(1:9), ]

informal=sum(who.inf$freq)/sum(who.sums$freq)
formal=sum(who.for$freq)/sum(who.sums$freq)

```


## Figure 1: If you had thoughts of suicide, whom would you trust to talk with about those thoughts? 
```{r}
#Create data frame containing options available for this questions
who = averages[1:30,]
##Remove the "who" header for the response option, and order the variables by decreasing values for graphing
who$Response = gsub("Who.",'',who$Response)
who = who[(order(-who$Percentage)),]
#Turn Proportion into Percentage
who$Percentage=(who$Percentage*100) 
#Round percentages to one decimal point
who$Percentage=round(who$Percentage,1)

p3=ggdotchart(who, y = "Percentage", x = "Response",
           color = "Suicide.Thoughts",     # Color by groups
           palette = c("I prefer not to say" = "bisque3",
                                "No." = "cadetblue4", "Yes." = "coral4"), # Custom color palette
           sorting = "ascending",  # Sort value in descending order
           rotate = TRUE,  # Rotate vertically
           #position=position_dodge(0.2),
           dot.size = 5,    # Large dot size'
          label = round(who$Percentage,1), repel=T, # Add dot labels
          label.rectangle=T,
          font.label = list(color = "Suicide.Thoughts", size = 6, face="bold", vjust=-1.5), 
          add="segments",
           ggtheme = theme_pubr()) +
  ylab("Percentage") +
  ggtitle("") +
    theme(plot.title = element_text(hjust = 0.5, size=14, face="bold")) +
   theme(axis.title.y=element_text(face="bold"),
            strip.text=element_text(face="bold", size=12),
            legend.title =element_text(size=10),
         axis.title.x =element_text(face="bold")) +
    theme_cleveland() 

p3


#save file as png for publication
ggsave(p3, file="who.png", height=5, width=6, dpi = 300)

```

## Figure 2: If you had thoughts of suicide, how would you bring it up with others?
```{r}
#Create data frame containing options available for this questions
how =averages[31:57,]
##Remove the "how" header for the response option, and order the variables by decreasing values for graphing
how$Response = gsub("How.",'',how$Response)
how = how[(order(-how$Percentage)),]
#Turn Proportion into Percentage
how$Percentage=(how$Percentage*100)
#Round percentages to one decimal point
how$Percentage=round(how$Percentage,1)

p4=ggdotchart(how, y = "Percentage", x = "Response",
           color = "Suicide.Thoughts",     # Color by groups
           palette = c("I prefer not to say" = "bisque3",
                                "No." = "cadetblue4", "Yes." = "coral4"), # Custom color palette
           sorting = "ascending",  # Sort value in descending order
           rotate = TRUE,  # Rotate vertically
           #position=position_dodge(0.2),
           dot.size = 5,    # Large dot size'
          label = round(how$Percentage,1), repel=T, # Add dot labels
          label.rectangle=T,
          font.label = list(color = "Suicide.Thoughts", size = 6, face="bold", vjust=-1.5), 
          add="segments",
           ggtheme = theme_pubr()) +
  ylab("Percentage") +
  ggtitle("") +
    theme(plot.title = element_text(hjust = 0.5, size=14, face="bold")) +
   theme(axis.title.y=element_text(face="bold"),
            strip.text=element_text(face="bold", size=12),
            legend.title =element_text(size=10),
         axis.title.x =element_text(face="bold")) +
    theme_cleveland() 

p4
#save file as png for publication
ggsave(p4, file="how.png", height=5, width=6, dpi = 300)

```

## Figure 3: If you had thoughts of suicide, what would you allow others to do to help you?
```{r}
#Create data frame containing options available for this questions
allow =averages[58:84,]
##Remove the "allow" header for the response option, and order the variables by decreasing values for graphing
allow$Response = gsub("Allow.",'',allow$Response)
allow = allow[(order(-allow$Percentage)),]
#Turn Proportion into Percentage
allow$Percentage=(allow$Percentage*100)
#Round percentages to one decimal point
allow$Percentage=round(allow$Percentage,1)

p5=ggdotchart(allow, y = "Percentage", x = "Response",
           color = "Suicide.Thoughts",     # Color by groups
           palette = c("I prefer not to say" = "bisque3",
                                "No." = "cadetblue4", "Yes." = "coral4"), # Custom color palette
           sorting = "ascending",  # Sort value in descending order
           rotate = TRUE,  # Rotate vertically
           #position=position_dodge(0.2),
           dot.size = 5,    # Large dot size'
          label = round(allow$Percentage,1), repel=T, # Add dot labels
          label.rectangle=T,
          font.label = list(color = "Suicide.Thoughts", size = 6, face="bold", vjust=-1.5), 
          add="segments",
           ggtheme = theme_pubr()) +
  ylab("Percentage") +
  ggtitle("") +
    theme(plot.title = element_text(hjust = 0.5, size=14, face="bold")) +
   theme(axis.title.y=element_text(face="bold"),
            strip.text=element_text(face="bold", size=12),
            legend.title =element_text(size=10),
         axis.title.x =element_text(face="bold")) +
    theme_cleveland() 

p5
#save file as png for publication
ggsave(p5, file="allow.png", height=5, width=6, dpi = 300)
```

## Multivariate Analysis

```{r}
#Allow Nothing
nothing.glm=glm(formula = Allow.Nothing ~ relevel(Suicide.Thoughts, ref="No.") + relevel(age.cat, ref="65-91") + relevel(Marital.Status, ref="Single ") + relevel(Status, ref="I am Currently Serving") + relevel(Branch, ref="Marine Corps"), data=complete.data, family="binomial")
summary(nothing.glm)
nothing.glm.OR=data.frame(exp(cbind(OR = coef(nothing.glm), confint(nothing.glm))))
nothing.glm.OR
```

```{r}
#2=Allow.Knowing
knowing.glm=glm(formula = Allow.Knowing ~ relevel(Suicide.Thoughts, ref="No.") + relevel(age.cat, ref="65-91") + relevel(Marital.Status, ref="Single ") + relevel(Status, ref="I am Currently Serving") + relevel(Branch, ref="Marine Corps"), data=complete.data, family="binomial")
summary(knowing.glm)
knowing.glm.OR=data.frame(exp(cbind(OR = coef(knowing.glm), confint(knowing.glm))))
knowing.glm.OR
```

```{r}
#3=Allow.Resourcing
resourcing.glm=glm(formula = Allow.Resourcing ~ relevel(Suicide.Thoughts, ref="No.") + relevel(age.cat, ref="65-91") + relevel(Marital.Status, ref="Single ") + relevel(Status, ref="I am Currently Serving") + relevel(Branch, ref="Marine Corps"), data=complete.data, family="binomial")
summary(resourcing.glm)
resourcing.glm.OR=data.frame(exp(cbind(OR = coef(resourcing.glm), confint(resourcing.glm))))
resourcing.glm.OR
```

```{r}
#4=Allow.MH.Treatment
MH.glm=glm(formula = Allow.MH.Treatment ~ relevel(Suicide.Thoughts, ref="No.") + relevel(age.cat, ref="65-91") + relevel(Marital.Status, ref="Single ") + relevel(Status, ref="I am Currently Serving") + relevel(Branch, ref="Marine Corps"), data=complete.data, family="binomial")
summary(MH.glm)
MH.glm.OR=data.frame(exp(cbind(OR = coef(MH.glm), confint(MH.glm))))
MH.glm.OR
```

```{r}
#5=Allow.Providers.In.My.House
prov.glm=glm(formula = Allow.Providers.In.My.House ~ relevel(Suicide.Thoughts, ref="No.") + relevel(age.cat, ref="65-91") + relevel(Marital.Status, ref="Single ") + relevel(Status, ref="I am Currently Serving") + relevel(Branch, ref="Marine Corps"), data=complete.data, family="binomial")
summary(prov.glm)
prov.glm.OR=data.frame(exp(cbind(OR = coef(prov.glm), confint(prov.glm))))
prov.glm.OR
```

```{r}
#6=Allow.App.To.Help
app.glm=glm(formula = Allow.App.To.Help ~ relevel(Suicide.Thoughts, ref="No.") + relevel(age.cat, ref="65-91") + relevel(Marital.Status, ref="Single ") + relevel(Status, ref="I am Currently Serving") + relevel(Branch, ref="Marine Corps"), data=complete.data, family="binomial")
summary(app.glm)
app.glm.OR=data.frame(exp(cbind(OR = coef(app.glm), confint(app.glm))))
app.glm.OR
```

```{r}
#7=Allow.Lock.Up.Guns
lockguns.glm=glm(formula = Allow.Lock.Up.Guns ~ relevel(Suicide.Thoughts, ref="No.") + relevel(age.cat, ref="65-91") + relevel(Marital.Status, ref="Single ") + relevel(Status, ref="I am Currently Serving") + relevel(Branch, ref="Marine Corps"), data=complete.data, family="binomial")
summary(lockguns.glm)
lockguns.glm.OR=data.frame(exp(cbind(OR = coef(lockguns.glm), confint(lockguns.glm))))
lockguns.glm.OR
```

```{r}
#8=Allow.Gun.Access.Notification
gunnot.glm=glm(formula = Allow.Gun.Access.Notification ~ relevel(Suicide.Thoughts, ref="No.") + relevel(age.cat, ref="65-91") + relevel(Marital.Status, ref="Single ") + relevel(Status, ref="I am Currently Serving") + relevel(Branch, ref="Marine Corps"), data=complete.data, family="binomial")
summary(gunnot.glm)
gunnot.glm.OR=data.frame(exp(cbind(OR = coef(gunnot.glm), confint(gunnot.glm))))
gunnot.glm.OR
```

```{r}
#9=Allow.Holding.Guns
gunhold.glm=glm(formula = Allow.Holding.Guns ~ relevel(Suicide.Thoughts, ref="No.") + relevel(age.cat, ref="65-91") + relevel(Marital.Status, ref="Single ") + relevel(Status, ref="I am Currently Serving") + relevel(Branch, ref="Marine Corps"), data=complete.data, family="binomial")
summary(gunhold.glm)
gunhold.glm.OR=data.frame(exp(cbind(OR = coef(gunhold.glm), confint(gunhold.glm))))
gunhold.glm.OR

```

```{r}
OR.combined=rbind(nothing.glm.OR, knowing.glm.OR, resourcing.glm.OR, MH.glm.OR, prov.glm.OR, app.glm.OR, lockguns.glm.OR, gunnot.glm.OR, gunhold.glm.OR)
OR.combined= OR.combined %>% mutate_if(is.numeric, round, digits=2)
OR.combined$stats <- paste(OR.combined$OR,"(",OR.combined$X2.5..,"-", OR.combined$X97.5.., ")")
write.csv(OR.combined, "OR.combined.csv", row.names=T)

```
